---
title: "Object Oriented Programming in R"
subtitle: "An Introduction to S3"
author: "Lukas Huwiler"
date: "November 25, 2020"
header-includes:
  - \usepackage{tikz}
  - \usepackage{tikz-uml}
  - \usetikzlibrary{arrows.meta}
  - \newcommand{\R}{\textsf{R}}
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

# A \textit{Very} Short Introduction to Object Oriented Programming

## Speaking Object Oriented

\begin{columns}[T]
\begin{column}{.4\textwidth}
\begin{minipage}[t]{\textwidth}
\includegraphics[width=\textwidth]{img/car.png}
\end{minipage}
\end{column}%
\hfill%
\begin{column}{.6\textwidth}
\begin{itemize} 
  \item It's a car $\rightarrow$ It's an \textit{instance} or \textit{object} of \textit{class} car
  \item It's a \textit{grey} car $\rightarrow$ The class car has an \textit{attribute} \texttt{color} (data)
  \item A car can \textit{drive} $\rightarrow$ The class car has a \textit{method} \texttt{drive} (behaviour)
\end{itemize}
\end{column}%
\end{columns}

## Class Hierarchies

\begin{columns}[T]
\begin{column}{.4\textwidth}
\begin{minipage}[t]{\textwidth}
\begin{tikzpicture}
  \umlclass[type=abstract,x=2]{Vehicle}{+ color: string}{+ drive(to)}
  \node[inner sep=0pt] (car) at (0,-4) {\includegraphics[width=.5\textwidth]{img/car.png}};
  \node[inner sep=0pt] (tuktuk) at (4,-4) {\includegraphics[width=.5\textwidth]{img/tuktuk.png}};
  \draw[-{Triangle[open,length=3mm,width=3mm]}, thick] (tuktuk.north) -- ([xshift=20pt,yshift=-2pt]Vehicle.south);
  \draw[-{Triangle[open,length=3mm,width=3mm]}, thick] (car.north) -- ([xshift=-20pt,yshift=-2pt]Vehicle.south);
\end{tikzpicture}
\end{minipage}
\end{column}%
\hfill%
\begin{column}{.6\textwidth}
\begin{itemize} 
  \item Cars and rickshaws are vehicles $\rightarrow$ Car and Rickshaw \textit{inherit} from Vehicle 
  \item Car and Rickshaw are child classes of Vehicle
  \item Vehicle is the parent class of Car and Rickshaw
\end{itemize}
\end{column}%
\end{columns}

# OOP in R

## It's a Mess!

\R has three (or even four) different class systems:

\begin{itemize}
  \item \textbf{S3:} oldest class system; very simple but elegant; comes with some limitations
  \item \textbf{S4:} more formal; adds multiple dispatch and type checks
  \item \textbf{R5:} Reference classes; feels similar to C++, Java, or Python; reference semantics
  \item \textbf{(R6):} the unofficial (but better?) implementation of R5
\end{itemize}
\vfill
I will focus on S3 in this talk\dots

# The S3 Class System

## Have You Ever Wondered?

```{r, echo=FALSE}
knitr::raw_latex("\\begin{tiny}\\begin{columns}[T]\n\\begin{column}{.65\\textwidth}\n")
car_model <- lm(mpg ~ hp, data = mtcars)
```
```{r}
summary(car_model)
```
```{r, echo=FALSE}
knitr::raw_latex("\\end{column}\n\\begin{column}{.4\\textwidth}\n")
```
```{r}
summary("It's all about cars")
summary(mtcars[1:2])
summary(table(mtcars$cyl))
summary(as.factor(mtcars$cyl))
```
```{r, echo=FALSE}
knitr::raw_latex("\\end{column}\n\\end{columns}\\end{tiny}")
```

The mechanism why summary works for (almost) everything is called \textit{method dispatch}

## How Method Dispatch Works

\scriptsize
```{r}
class(mtcars)
summary
methods(summary)[1:16]
```

## Create Your Own Generic Method

\tiny
```{r}
# Define methods
drive <- function(x, location) UseMethod("drive", x)

drive.Car <- function(x, location) {
  cat("drive car...\n")
  x$location <- location
  x$gas <- pmax(x$gas - 10, 0)
  x
}

drive.Rickshaw <- function(x, location) {
  cat("drive rickshaw...\n")
  x$location <- location
  x
}

# Instantiate objects
car <- structure(list(color = "grey", location = "Zurich", gas = 100), 
                 class = "Car")
rickshaw <- list(color = "blue", location = "Munich")
class(rickshaw) <- "Rickshaw"
```
```{r}
# Call generic methods
car <- drive(car, "Lucerne")
rickshaw <- drive(rickshaw, "Lucerne")
```

## Class Hierarchies in S3

\tiny
```{r}
# Make the objects inheriting from Vehicle (don't show this to a Java developer)
class(car) <- c("Car", "Vehicle")
class(rickshaw) <- c("Rickshaw", "Vehicle")
rm(drive.Rickshaw)

drive.Vehicle <- function(x, location) {
  cat("Change location...\n")
  x$location <- location
  x
}

drive.Car <- function(x, location) {
  x <- NextMethod()
  cat("Consume gasoline...\n")
  x$gas <- pmax(x$gas - 10, 0)
  x
}
```
```{r}
car <- drive(car, "Lucerne")
rickshaw <- drive(rickshaw, "Lucerne")
```

## Some Additional Points

\scriptsize
```{r}
# You can define a default method as a fallback option:
drive.default <- function(x, location) {
  sprintf("An instance of class %s cannot drive...", class(x))
}

drive(2, "Lucerne")

# The package sloop offers some nice functions for a better understanding:
sloop::s3_dispatch(drive(car, "Zurich"))

# It's a good idea to write constructors for your S3 class:
Car <- function(location, color = "black", gas_level = 100) {
  car <- list(color = color, location = location, gas = gas)
  class(car) <- c("Car", "Vehicle")
  car
}
```

##

\Huge
Questions?
