---
title: "Object Oriented Programming in R"
subtitle: "An Introduction to S3"
author: "Lukas Huwiler"
date: "November 25, 2020"
header-includes:
  - \usepackage{tikz}
  - \usepackage{tikz-uml}
  - \usetikzlibrary{arrows.meta}
  - \newcommand{\R}{\textsf{R}}
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---


# A \textit{Very} Short Introduction to Object Oriented Programming

## Object Oriented Programming (OOP) is \dots

- \dots \textit{the} dominant programming paradigm since the 90s
- \dots a way to abstract and code
- \dots all about keeping data and behaviour together

## Speaking Object Oriented

\begin{columns}[T]
\begin{column}{.4\textwidth}
\begin{minipage}[t]{\textwidth}
\includegraphics[width=\textwidth]{img/car.png}
\end{minipage}
\end{column}%
\hfill%
\begin{column}{.6\textwidth}
\begin{itemize} 
  \item It's a car $\rightarrow$ It's an \textit{instance} or \textit{object} of \textit{class} car
  \item It's a \textit{grey} car $\rightarrow$ The class car has an \textit{attribute} \texttt{colour} (data)
  \item A car can \textit{drive} $\rightarrow$ The class car has a \textit{method} \texttt{drive} (behaviour)
\end{itemize}
\end{column}%
\end{columns}

## Class Hierarchies

\begin{columns}[T]
\begin{column}{.4\textwidth}
\begin{minipage}[t]{\textwidth}
\begin{tikzpicture}
  \umlclass[type=abstract,x=2]{Vehicle}{+ color: string}{+ drive(to)}
  \node[inner sep=0pt] (car) at (0,-4) {\includegraphics[width=.5\textwidth]{img/car.png}};
  \node[inner sep=0pt] (tuktuk) at (4,-4) {\includegraphics[width=.5\textwidth]{img/tuktuk.png}};
  \draw[-{Triangle[open,length=3mm,width=3mm]}, thick] (tuktuk.north) -- ([xshift=20pt,yshift=-2pt]Vehicle.south);
  \draw[-{Triangle[open,length=3mm,width=3mm]}, thick] (car.north) -- ([xshift=-20pt,yshift=-2pt]Vehicle.south);
\end{tikzpicture}
\end{minipage}
\end{column}%
\hfill%
\begin{column}{.6\textwidth}
\begin{itemize} 
  \item Cars and rickshaws are vehicles $\rightarrow$ Cars and rickshaws \textit{inherit} from Vehicle 
  \item You will never see a vehicle (only types of vehicles) $\rightarrow$ Vehicle is an \textit{abstract class}
  \item \dots and this is not just a diagram, it's an UML class diagram $\rightarrow$ OOP even has it's own graphical representation language
\end{itemize}
\end{column}%
\end{columns}



# OOP in R

## It's a Mess!

\R has three (or even four) different class systems:

\begin{columns}[T]
\begin{column}{0.6\textwidth}
\begin{itemize}
  \item \textbf{S3:} oldest class system; very simple but elegant; comes with some limitations
  \item \textbf{S4:} more formal; adds multiple dispatch and type checks
  \item \textbf{R5:} Reference classes; feels similar to C++, Java, or Python; reference semantics
  \item \textbf{(R6):} the unofficial (but better?) implementation of R5
\end{itemize}
\end{column}
\hfill
\begin{column}{0.4\textwidth}

\end{column}
\end{columns}
\vfill
We will focus on S3 for this talk\dots

## It's a Mess!

\R has three (or even four) different class systems:

\begin{columns}[T]
\begin{column}{0.6\textwidth}
\begin{itemize}
  \item \textbf{S3:} oldest class system; very simple but elegant; comes with some limitations
  \item \textbf{S4:} more formal; adds multiple dispatch and type checks
  \item \textbf{R5:} Reference classes; feels similar to C++, Java, or Python; reference semantics
  \item \textbf{(R6):} the unofficial (but better?) implementation of R5
\end{itemize}
\end{column}
\hfill
\begin{column}{0.4\textwidth}
\begin{tikzpicture}
  \node[inner sep=0pt] (noop) at (0,0) {\textit{not really object oriented}};
  \node[inner sep=0pt] (oop) at (0,-4) {\textit{object oriented}};
  \draw[<->, thick] ([yshift=-2pt]noop.south) -- ([yshift=2pt]oop.north);
\end{tikzpicture}
\end{column}
\end{columns}
\vfill
We will focus on S3 for this talk\dots

# The S3 Class System

## Have You Ever Wondered?

```{r, echo=FALSE}
knitr::raw_latex("\\begin{tiny}\\begin{columns}[T]\n\\begin{column}{.65\\textwidth}\n")
car_model <- lm(mpg ~ hp, data = mtcars)
```
```{r}
summary(car_model)
```
```{r, echo=FALSE}
knitr::raw_latex("\\end{column}\n\\begin{column}{.4\\textwidth}\n")
```
```{r}
summary("It's all about cars")
summary(mtcars[1:2])
summary(table(mtcars$cyl))
summary(as.factor(mtcars$cyl))
```
```{r, echo=FALSE}
knitr::raw_latex("\\end{column}\n\\end{columns}\\end{tiny}")
```

The mechanism why summary works for (almost) everything is called \textit{method dispatch}

## How Method Dispatch Works

\scriptsize
```{r}
class(mtcars)
attr(car_model, "class")
summary
methods(summary)[1:16]
```

## Create Your Own Generic Method

\scriptsize
```{r}
# Define methods
drive <- function(x) {
  UseMethod("drive", x)
}

drive.Car <- function(x) {
  print("brumm, brumm")
}

drive.Rikshaw <- function(x) {
  print("tuk, tuk")
}

# Instantiate objects
car <- structure(list(color = "grey"), class = "Car")
tuktuk <- structure(list(color = "blue"), class = "Rikshaw")

# Call generic methods
drive(car)
drive(tuktuk)
```

## Class Hierarchies in S3

\scriptsize
```{r}
paint <- function(x, color) {
  UseMethod("paint")
}

paint.Vehicle <- function(x, color) {
  x$color <- color
  x
}
# Java developers start sweating here
class(car) <- c("Car", "Vehicle")
paint(car, color = "pink")
```

## Class Hierarchies in S3

\scriptsize
```{r}
paint.Car <- function(x, color) {
  painted_car <- NextMethod() # print.Vehicle method does the basics
  print("Additionally we'll change the seat's color")
  painted_car
}

paint(car, "black")

paint.default <- function(x, color) {
  sprintf("How am I supposed to paint %s?", as.character(x))
}

paint(2, "red")
```

## S3 in the Field

It's a good idea to write constructors for your S3 class:

\scriptsize
```{r}
new_Car <- function(color) {
  structure(list(color = color), class = c("Car", "Vehicle"))
}

validate_Car <- function(x) {
  if (!inherits(x$color, "character")) {
    stop("color has to be of class character", .call = TRUE) 
  }
  x
}

Car <- function(color = "black") {
  car <- new_Car(color)
  validate_Car(car)
  car
}
```

## The End

\Huge
Questions?
